---
title: "Crane Mycobiomes"
author: "Chauncey Gadek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    highlight: tango
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=F}
pacman::p_load(
tidyverse,
dplyr,
vegan,
ggplot2,
viridis,
cowplot,
randomcoloR,
plotrix 
  
)



df <- read.csv("data/morphotype_seq.csv", stringsAsFactors = T)

# make column for higher order ID
df$higher_ID <- NA
df$Euro <- NA
for(i in 1:nrow(df)){
  df$higher_ID[i] <- sub(" .*", "", df$Tentative.ID[i])
  if(df$higher_ID[i] %in% c("Aspergillus", "Penicillium")){
    df$Euro[i] <- "Euro"
  }else{
    df$Euro[i] <- "Not"
  }
}
pathogenic <- c("Apergillus", "Alternaria", "Aureobasidium", "Cladosporium","Chaetomium", "Lichtheimia", "Mucor", "Papiliotrema", "Penicillium", "Rhizopus", "Rhodotorula", "Uncultured")
df<-df%>% mutate(path=ifelse(higher_ID %in% pathogenic, "p","np"))
df$higher_ID <- as.factor(df$higher_ID)
df$Euro <- as.factor(df$Euro)
df$path <- as.factor(df$path)

proportions <- df %>%
  group_by(subspecies, Tentative.ID)%>% 
  summarize(n = length(Tentative.ID)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) 

n_lessers <- df[df$subspecies=="Antigone canadensis canadensis",]
n_lessers <- length(unique(n_lessers$MSB_ID))

n_greaters <- df[df$subspecies=="Antigone canadensis tabida",]
n_greaters <- length(unique(n_greaters$MSB_ID))

```

# Summaries by subspecies
## number of tentative ID taxa per subspecies
```{r, echo=F}
df %>%
  group_by(subspecies, Tentative_ID_short)%>% 
  summarise(no_rows = length(unique(subspecies)))%>%
  group_by(subspecies)%>% 
  summarise(no_rows = length(subspecies))%>%
  ggplot(., aes(subspecies, no_rows))+
  geom_col()+
  ggtitle("OTUs per subspecies")+
  scale_fill_manual(values="grey99")+
  theme_classic()+
  scale_x_discrete(labels=c("lesser", "greater"))+
  labs(x="subspecies", y="number unique OTUs")+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.position = "none",
        axis.text=element_text(size=14),
        axis.title=element_text(size=14,face="bold"))


```


## number of tentative ID taxa per subspecies corrected for sample sizes
```{r, echo=F}

t<-df %>%
  group_by(subspecies, Tentative.ID)%>% 
  summarise(no_rows = length(unique(subspecies)))%>%
  group_by(subspecies)%>% 
  summarise(no_rows = length(subspecies))

v<- t$no_rows[1]/n_lessers
c <- t$no_rows[2]/n_greaters

t$corr <-NA
t$corr[1] <-v
t$corr[2] <-c 
t%>%
  ggplot(., aes(subspecies, corr))+
  geom_col()+
  ggtitle("OTUs per subspecies")+
  scale_fill_manual(values="grey62")+
  theme_classic()+
  scale_x_discrete(labels=c("lesser", "greater"))+
  labs(x="subspecies", y="unique OTUs/indivudal")+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.position = "none",
        axis.text=element_text(size=14),
        axis.title=element_text(size=14,face="bold"))


```


```{r, echo=F}
df %>%
  group_by(subspecies, MSB_ID)%>% 
  summarize(n = length(unique(Tentative.ID)))%>%
  summarise(mean= mean(n),
            se = std.error(n))
```
Essentially the same number of isolated fungal taxa from each subspecies. Suggesting that lessers have higher diversity given they have nearly 1/5 the number of individuals as greaters.


```{r, echo=F}
df %>%
  group_by(subspecies, MSB_ID)%>% 
  summarize(n = length(unique(Tentative.ID)))%>%
   summarise(mean= mean(n),
            se = std.error(n))%>%
ggplot(., aes(subspecies, n, fill=subspecies))+
  geom_col()
```

```{r, echo=F}
df %>%
  group_by(subspecies, MSB_ID)%>% 
  summarize(n = length(unique(higher_ID)))%>%
  summarise(mean= mean(n),
            se = std.error(n))
```
Essentially the same number of isolated fungal taxa from each subspecies at higher taxonomic level.


## number of higher ID taxa per subspecies
```{r, echo=F}
df %>%
  group_by(subspecies)%>% 
  summarize(n = length(unique(df$Tentative_ID_short)))
```


## number of higher ID taxa per subspecies
```{r, echo=F}
df %>%
  group_by(subspecies)%>% 
  summarize(n = length(unique(higher_ID)))
```

# Plot morphotype proportions
## Specific ID
```{r, echo=F}
#set up color palettes
n <- length(levels(df$Tentative.ID))
palette <- distinctColorPalette(n)

df %>%
  group_by(subspecies, Tentative.ID)%>% 
  summarize(n = length(Tentative.ID)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
ggplot(., aes(fill = Tentative.ID,
                      y = proportion, x = subspecies))+
  geom_bar(position = "fill", stat = "identity")+
  ggtitle("Tentative Culture ID")+
  scale_fill_manual(values=palette)+
  theme_classic(base_size = 12)+
  scale_x_discrete(labels=c("lesser", "greater"))+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title = element_text(size = 3), 
               legend.text = element_text(size = 3))

```

# Specific ID 
but cleaned because pf weird labeling issues
```{r, echo=F}
#set up color palettes
n <- length(levels(df$Tentative_ID_short))
palette <- distinctColorPalette(n)

df %>%
  group_by(subspecies, Tentative_ID_short)%>% 
  summarize(n = length(Tentative_ID_short)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
ggplot(., aes(fill = Tentative_ID_short,
                      y = proportion, x = subspecies))+
  geom_bar(position = "fill", stat = "identity")+
  ggtitle("Tentative Culture ID")+
  scale_fill_manual(values=palette)+
  theme_classic(base_size = 12)+
  scale_x_discrete(labels=c("lesser", "greater"))+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title = element_text(size = 3), 
               legend.text = element_text(size = 3))

```
## Higher ID
```{r, echo=F}
n <- length(levels(df$higher_ID))
palette <- distinctColorPalette(n)

df %>%
  group_by(subspecies, higher_ID)%>% 
  summarize(n = length(higher_ID)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
ggplot(., aes(fill = higher_ID,
                      y = proportion, x = subspecies))+
geom_bar(position = "fill", stat = "identity")+
  ggtitle("Tentative Culture ID")+
  scale_fill_viridis_d(option="turbo")+
  #scale_fill_manual(values=palette)+
  theme_classic(base_size = 12)+
  scale_x_discrete(labels=c("lesser", "greater"))+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title = element_text(size = 3), 
               legend.text = element_text(size = 3))

```


## family ID
```{r, echo=F}
n <- length(levels(df$family))
palette <- distinctColorPalette(n)

df %>%
  group_by(subspecies, family)%>% 
  summarize(n = length(family)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
ggplot(., aes(fill = family,
                      y = proportion, x = subspecies))+
geom_bar(position = "fill", stat = "identity")+
  ggtitle("Family")+
  scale_fill_viridis_d(option="turbo")+
  #scale_fill_manual(values=palette)+
  theme_classic(base_size = 12)+
  scale_x_discrete(labels=c("lesser", "greater"))+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title=element_blank(), 
               legend.text = element_text(size = 3))+
  guides(fill = guide_legend(ncol = 2, byrow = T))

```


## order ID
```{r, echo=F}
n <- length(levels(df$order))
palette <- distinctColorPalette(n)

p<-df %>%
  group_by(subspecies, order)%>% 
  summarize(n = length(order)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
  arrange(desc(proportion), group_by=subspecies)%>%
ggplot(., aes(fill = fct_reorder(order,proportion),
                      y = proportion, x = subspecies))+
geom_bar(position = "fill", stat = "identity", color="black")+
  ggtitle("Order")+
scale_fill_manual(values=c("grey44", "sienna3", "goldenrod3", "bisque3", "grey90", "azure3", "thistle3", "lemonchiffon3", "steelblue", "cornsilk2", "peachpuff2", "lightblue3", "navajowhite3", "mistyrose2", "plum4"))+
  scale_x_discrete(labels=c("lesser", "greater"))+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title=element_blank(), 
               legend.text = element_text(size = 3))+
  guides(fill = guide_legend(ncol = 2, byrow = T))+
  theme_cowplot()
p


ggsave(p, file="figures/Proportion_orders.pdf", width=14.6, height=7.2)
```

## Just Apsergillus OTUs
```{r, echo=F}


p<-df %>%
  filter(higher_ID=="Aspergillus")%>%
  group_by(subspecies, Tentative_ID_short)%>% 
  summarize(n = length(Tentative_ID_short)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
ggplot(., aes(fill = fct_reorder(Tentative_ID_short,proportion),
                      y = proportion, x = subspecies))+
geom_bar(position = "fill", stat = "identity", color="black")+
  ggtitle("Aspergillus OTU")+
scale_fill_manual(values=c("grey44", "sienna3", "goldenrod3", "bisque3", "grey90", "azure3", "thistle3", "lemonchiffon3"))+
  scale_x_discrete(labels=c("lesser", "greater"))+
  labs(x="", y="proportion unique OTUs")+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title=element_blank(), 
        legend.text = element_text(size = 4))+
  theme_cowplot()
p

ggsave(p, file="figures/Proportion_Aspergillus.pdf", width=7.7, height=3.6)
```

## Just Penicillium OTUs
```{r, echo=F}

p<-df %>%
  filter(higher_ID=="Penicillium")%>%
  group_by(subspecies, Tentative_ID_short)%>% 
  summarize(n = length(Tentative_ID_short)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
ggplot(., aes(fill = fct_reorder(Tentative_ID_short,proportion),
                      y = proportion, x = subspecies))+
geom_bar(position = "fill", stat = "identity", color="black")+
  ggtitle("Penicillium OTU")+
 scale_fill_manual(values=c("grey44", "sienna3", "goldenrod3", "bisque3", "grey90", "azure3"))+
  scale_x_discrete(labels=c("lesser", "greater"))+
  labs(x="", y="proportion unique OTUs")+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title=element_blank(), 
               legend.text = element_text(size = 3))+
  theme_cowplot()
p

ggsave(p, file="figures/Proportion_Penicillium.pdf", width=7.7, height=3.6)
```
## For  Penicillium and Aspergillus combined
 
```{R}
p<- df %>%
 # filter(higher_ID==c("Penicillium", "Aspergillus"))%>%
  group_by(subspecies, Euro)%>% 
  summarize(n = length(Euro)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
ggplot(., aes(fill = Euro,
                      y = proportion, x = subspecies))+
geom_bar(position = "fill", stat = "identity", color="black")+
  ggtitle("Penicillium OTU")+
  scale_fill_manual(values=c( "sienna3", "grey44", "goldenrod3", "bisque3", "grey90", "azure2"))+
  scale_x_discrete(labels=c("lesser", "greater"))+
  labs(x="", y="proportion Eurotiales OTUs")+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title=element_blank(), 
               legend.text = element_text(size = 3))+
  theme_cowplot()

p


ggsave(p, file="figures/Proportion_Eurotiales.pdf", width=5.6, height=3.6)
```

Chi-squared test for Eurotailes between subpecies
```{R}
chisq.test(df$subspecies, df$Euro, simulate.p.value = T)

```
Next I am doing a literature search by OTU higher order taxon name to determine if I can include them in opportunistic infection category or not...

Oppurtunistic infection in birds:
1. Apergillus
2. Alternaria
3. Aureobasidium bird feathers
4. Cladosporium
5. Chaetomium bird feathers showed higher depredation on individuals with feather growth of this fungi https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.3582
6. 	Lichtheimia 
7. Mucor
8. Papiliotrema 
9. Penicillium
10. Rhizopus
11. Rhodotorula
12. Uncultured Dothidiomycetes

```{r}

pathogenic <- c("Apergillus", "Alternaria", "Aureobasidium", "Cladosporium","Chaetomium", "Lichtheimia", "Mucor", "Papiliotrema", "Penicillium", "Rhizopus", "Rhodotorula", "Uncultured")

  
p<-df%>%
  group_by(subspecies, path)%>% 
  mutate(path=ifelse(higher_ID %in% pathogenic, "p","np"))%>%
  summarize(n = length(path)) %>%
  ungroup %>% 
  group_by(subspecies) %>% 
  mutate(proportion = n / sum(n)) %>%
  ggplot(., aes(fill = path,
                      y = n, x = subspecies))+
geom_bar(position = "fill", stat = "identity", color="black")+
  ggtitle("Opportunistic Pathogen OTUs")+
  scale_fill_manual(values=c("bisque2", "sienna3"), labels=c( "not \n pathogenic", "opportunistic \n pathogen"))+
  scale_x_discrete(labels=c("lesser", "greater"))+
  labs(x="", y="proportion pathogen OTUs")+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title = element_text(size = 0), 
               legend.text = element_text(size = 8))+
  theme_cowplot()
p
ggsave(p, file="figures/Proportion_oppurtunistic_pathogen.pdf", width=5.6, height=3.6)
```



```{r}
chisq.test(df$subspecies, df$Euro, simulate.p.value = T)
```

## MDS
```{r}

library(magrittr)
library(dplyr)
library(ggpubr)
row <- df %>%
  group_by(subspecies, MSB_ID, Tentative.ID) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = Tentative.ID, values_from = n) 

mds <- df %>%
   unite(subid, c("subspecies", "MSB_ID"))%>%
  group_by(subid,higher_ID) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = higher_ID, values_from = n) %>%
  replace(is.na(.), 0)%>%
  column_to_rownames(., var="subid")%>%
  as.matrix()
#   dist() %>%          
#   cmdscale() %>%
#   as_tibble()
# colnames(mds) <- c("Dim.1", "Dim.2")


c.myco.NMDS=metaMDS(mds, # Our community-by-species matrix
                     k=2) # The number of reduced dimensions




mds$subpecies <- row$subspecies
# Plot MDS

ggplot(mds, aes(Dim.1, Dim.2, fill=subpecies))+
  geom_point(shape=21)



```
## NMDS

```{r}
mds <- df %>%
   unite(subid, c("subspecies", "MSB_ID"))%>%
  group_by(subid,order) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = order, values_from = n) %>%
  replace(is.na(.), 0)%>%
  column_to_rownames(., var="subid")%>%
  as.matrix()


c.myco.NMDS=metaMDS(mds, # Our community-by-species matrix
                     k=2) # The number of reduced dimensions



stressplot(c.myco.NMDS)
```

# By age
## lesser subspecies
### Higher ID
```{r, echo=F}
n <- length(levels(df$order))
palette <- distinctColorPalette(n)

df %>%
  filter(subspecies %in% "Antigone canadensis canadensis")%>%
  mutate(age2=ifelse(age %in% c("a"), "a","i") )%>%
  group_by(age2, order,)%>% 
  summarize(n = length(order)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
ggplot(., aes(fill = order,
                      y = proportion, x = age2))+
geom_bar(position = "fill", stat = "identity")+
  ggtitle("Order Lessers by Age")+
  scale_fill_viridis_d(option="plasma")+
  #scale_fill_manual(values=palette)+
  theme_classic(base_size = 12)+
  scale_x_discrete(labels=c("adult", "immature"))+
  labs(x="age", y = "proportion OTU")+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title = element_text(size = 0), 
               legend.text = element_text(size = 3))

```

### Just Apsergillus OTUs
```{r, echo=F}
n <- length(levels(df$higher_ID))
palette <- distinctColorPalette(n)

df %>%
  filter(higher_ID=="Aspergillus",
         #sex %in% c("m", "f"),
         subspecies %in% "Antigone canadensis canadensis")%>%
  mutate(age2=ifelse(age %in% c("a"), "a","i") )%>%
  group_by(age2, Tentative_ID_short)%>% 
  summarize(n = length(Tentative_ID_short)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
ggplot(., aes(fill = Tentative_ID_short,
                      y = proportion, x = age2))+
geom_bar(position = "fill", stat = "identity")+
  ggtitle("Aspergillus OTU")+
  #scale_fill_manual(values=palette)+
  scale_fill_viridis_d(option="plasma")+
  theme_classic(base_size = 12)+
  scale_x_discrete(labels=c("adult", "immature"))+
  labs(x="age", y="proportion unique OTUs")+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title = element_text(size = 0), 
               legend.text = element_text(size = 4))

```
Interesting no immature lesser birds have any Aspergillus OTUs.

### Putative pathogens
```{r}
pathogenic <- c("Apergillus", "Alternaria", "Aureobasidium", "Cladosporium","Chaetomium", "Lichtheimia", "Mucor", "Papiliotrema", "Penicillium", "Rhizopus", "Rhodotorula", "Uncultured")

df%>%
  filter(subspecies %in% "Antigone canadensis canadensis")%>%
  mutate(path=ifelse(higher_ID %in% pathogenic, "p","np"),
         age2=ifelse(age %in% c("a"), "a","i"))%>%
  group_by(age2, path)%>% 
  summarize(n = length(path)) %>%
  #ungroup %>% group_by(state, gender) %>% 
  mutate(proportion = n / sum(n)) %>%
  ggplot(., aes(fill = path,
                      y = n, x = age2))+
geom_bar(position = "fill", stat = "identity")+
  ggtitle("Opportunistic Pathogen OTUs in Lessers" )+
  scale_fill_manual(values=c("goldenrod3", "dodgerblue3"), labels=c( "not \n pathenogenic", "oppurtunistic \n pathogen"))+
  theme_classic(base_size = 12)+
  scale_x_discrete(labels=c("adult", "immature"))+
  labs(x="age", y="proportion unique OTUs")+
  guides(shape = guide_legend(override.aes = list(size = 0.5)))+
  guides(fill= guide_legend(override.aes = list(size = 0.5)))+
  theme(legend.title = element_text(size = 0), 
               legend.text = element_text(size = 5))

chisq.test(df$subspecies, df$path, simulate.p.value = T)
```